---
title: "Test"
output: html_document
date: "2024-09-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
if(!is.null(dev.list())) dev.off() # Clear plots in case these are present
rm(list=ls()) # Remove all variables that are in the current workspace
```

##Data preparation

First, the data was imported and sampled as asked to. 

```{r}
library(readr)
train <- read_csv("train.csv")
View(train)
```

```{r}
head(train)
```
We need some information about the data to start making decisions about how deal with the problem.


```{r}
summary(train)
```
In the summary, it can be notice that in the variable "pctsomecol18_24" shows a great percentage of missing values. According with the information of the project, the variable corresponds with "Percent of county residents ages 18-24 highest education attained: some college" so it should be decided what to do with the variable.

```{r}
str(train)
```
This checks the type of each variable 
```{r}
# Some plots are done:
boxplot(train$percentmarried)
boxplot(train$pctmarriedhouseholds)
```

```{r}
varout <- summary(train$pctmarriedhouseholds)

# Interquartile range calculation:
iqr <- varout[5] - varout[2]

umout <- varout[5] + 1.5*iqr # Upper extreme for mild outliers
usout <- varout[5] + 3*iqr # Upper extreme for extreme outliersm (iqr = inter quartile range)
boxplot(train$percentmarried, horizontal = TRUE)
boxplot(train$pctmarriedhouseholds, horizontal = TRUE)
abline(v = umout, col = "orange", lwd = 2)
abline(v = usout, col = "red", lwd = 2)
```

```{r}
plot(train[,c(14,32)])
```

```{r}
cor(train[,1:8])
```

